# Code-a-thon Project Template

###

This project is intended to help you apply your skills in R in a way that would be conducive for collaborating or creating reproducible analyses.The project I'm working on is trying to correlate hospitalization rate spikes to heat waves, and PM2.5 spikes affect on smoke and asthma hospitalization rates.

cvcv
## Data Used

Summer hazards excel file was downloaded on 6/28/2026 from the Washington State Dept. of Health Seasonal Hazards Dashboard. 

Url: https://doh.wa.gov/public-health-provider-resources/healthcare-professions-and-facilities/data-exchange/syndromic-surveillance-rhino/seasonal-hazards-dashboard

Data import into R:
```{r}
library(tidyverse)
library(rio)
library(ggplot2)
library(maps)
library(lmtest)

rhino <- import("C:/Users/mkett/Desktop/DaSEH project/rhino_sh_summer_hazards_downloadable.csv")
glimpse(rhino)
#load washington county map
wa_county <- map_data("county") |> filter(region == "washington")

```

The data contains hospitalization data visits by type of hazard (asthma, boating, drowning, heat, and smoke), between May 1st and October 31st for years 2022-present. 
It has individual county data as well as statewide summaries. Key independent variables are max temp and PM2.5 particulate and how it affects hospitalizations for those hazards. 


# Data Cleaning/Wrangling
Based on a glimpse of the data, I changed Date variable from character to date type, also I removed the date-time Stamp variable as that column isn't needed, just shows the time downloaded. 

# Joining Map data to get county maps for 2025 year


```{r}
# adjusting Date from character to date type and removing data time stamp column not needed.
#rhino <- rhino |> select(-'Date-Time Stamp')
rhino$Date <-mdy(rhino$Date)
#check data again
glimpse(rhino)

#county only multiple years
rhino_county <- rhino |> filter(geography == "County") |> 
  mutate(location = tolower(str_remove(location, " County")))|>
  inner_join(wa_county, by = join_by(location==subregion))
#summary map

#2025 county data
rhino_2025_co <- rhino_county |> filter(year==2025)
# county data is ready for subset to make maps for heat, smoke and asthma


```


# Data Visualization
Create maps with max temp and smoke for 2025. 

Also creat maps were created for each hazard of interest(heat, smoke, and asthma) to find counties with the highest hospitalization percentages. Some counties don't have temperature or PM2.5 data so future work would be to gather that data. Otherwise we will use counties that contain that data. 

```{r}
#map 2025 highest PM2.5 and max temp
rhino_2025_maxtemp <- rhino_2025_co |>
  group_by(location) |> summarize(max_temp_degF = max(max_temp_degF)) |>
  full_join(wa_county, by = join_by(location==subregion))

rhino_2025_pm2.5 <- rhino_2025_co |>
  group_by(location) |> summarize(max_pm2.5 = max(max_pm2.5)) |>
  full_join(wa_county, by = join_by(location==subregion))

#map 2025 sum of heat hospitalizations
rhino_2025_sum_heat <- rhino_2025_co |>
  filter(hazard == "heat") |>
  group_by(location) |> summarize(max_pct_ed_visits = max(pct_ed_visits)) |>
  full_join(wa_county, by = join_by(location==subregion)) 
#2026 sum of smoke hospitalizations
rhino_2025_sum_smoke <- rhino_2025_co |>
  filter(hazard == "smoke") |>
  group_by(location) |> summarize(max_pct_ed_visits = max(pct_ed_visits)) |>
  full_join(wa_county, by = join_by(location==subregion))
#2026 sum of asthma hospitalizations
rhino_2025_sum_asthma <- rhino_2025_co |>
  filter(hazard == "asthma") |>
  group_by(location) |> summarize(max_pct_ed_visits = max(pct_ed_visits)) |>
  full_join(wa_county, by = join_by(location==subregion))


#plot this
plot_2025_maxtemp <-  ggplot() + 
  geom_polygon(data = rhino_2025_maxtemp, 
               aes(x = long, y = lat, group = group, fill = max_temp_degF), color="red") +
  #geom_point(data = wa_cities, aes(x = long, y = lat), color = "red") +
  labs(
    title = "Washington max temps 2025 (deg F)",
    x = "longitude", y = "latitude") +
  coord_fixed(1.3)
plot_2025_maxtemp

plot_2025_pm2.5 <-  ggplot() + 
  geom_polygon(data = rhino_2025_pm2.5, 
               aes(x = long, y = lat, group = group, fill = max_pm2.5), color="red") +
  #geom_point(data = wa_cities, aes(x = long, y = lat), color = "red") +
  labs(
    title = "Washington max PM2.5 2025",
    x = "longitude", y = "latitude") +
  coord_fixed(1.3)
plot_2025_pm2.5

plot_2025_heat <-
  ggplot() + 
  geom_polygon(data = rhino_2025_sum_heat, 
               aes(x = long, y = lat, group = group, fill = max_pct_ed_visits), color="red") +
  #geom_point(data = wa_cities, aes(x = long, y = lat), color = "red") +
  labs(
    title = "Washington max heat related emergency visits, 2025",
    x = "longitude", y = "latitude") +
  coord_fixed(1.3)
plot_2025_heat

plot_2025_smoke <- 
  ggplot() + 
  geom_polygon(data = rhino_2025_sum_smoke, 
               aes(x = long, y = lat, group = group, fill = max_pct_ed_visits), color="red") +
  #geom_point(data = wa_cities, aes(x = long, y = lat), color = "red") +
  labs(
    title = "Washington max smoke related emergency visits, 2025",
    x = "longitude", y = "latitude") +
  coord_fixed(1.3)
plot_2025_smoke

plot_2025_asthma <- 
  ggplot() + 
  geom_polygon(data = rhino_2025_sum_asthma, 
               aes(x = long, y = lat, group = group, fill = max_pct_ed_visits), color="red") +
  #geom_point(data = wa_cities, aes(x = long, y = lat), color = "red") +
  labs(
    title = "Washington max asthma related emergency visits, 2025",
    x = "longitude", y = "latitude") +
  coord_fixed(1.3)
plot_2025_asthma


```


# Data Analysis

Based on the maps above identifying higher values of hospitalizations, will use the time series for that county to do a plot and do a predictive test for temp on hospitalizations for adams county. 

```{r}
rhino_2025_adams <- rhino |> filter(geography == "County") |> 
  mutate(location = tolower(str_remove(location, " County")))|>
  filter(location == "adams") |> filter(year == 2025) |> filter(hazard == "heat")
plot_ts_2025_adams <-
  ggplot(data = rhino_2025_adams, 
         aes(x=Date, y=max_temp_degF)) +  geom_line() + scale_x_date(date_breaks = "1 month") +
        labs(
          title = "Daily temperature in Adams county summer 2025", 
          y = "Temperature (degrees F)"
        )
plot_ts_2025_adams

#try plotting hospital rates and temp on same graph
adams_2025.long <-rhino_2025_adams |>
    select(Date, max_temp_degF, pct_ed_visits) |>
    pivot_longer(-Date, names_to = "variable", values_to = "value")
ggplot(adams_2025.long, aes(Date, value, colour = variable)) + geom_line()

#granger causality test
grangertest(pct_ed_visits~max_temp_degF, order = 2, data = rhino_2025_adams)
# p-value is 0.18 with 2 lags, try adding more data multiple years

rhino_all_adams <- rhino |> filter(geography == "County") |> 
  mutate(location = tolower(str_remove(location, " County")))|>
  filter(location == "adams") |> filter(hazard == "heat")
adams_all.long <-rhino_all_adams |>
  select(Date, max_temp_degF, pct_ed_visits) |>
  pivot_longer(-Date, names_to = "variable", values_to = "value")
ggplot(adams_all.long, aes(Date, value, colour = variable)) + geom_line()
grangertest(pct_ed_visits~max_temp_degF, order = 1, data = rhino_all_adams)
#granger test is significant with all years of data
grangertest(max_temp_degF~pct_ed_visits, order = 1, data = rhino_all_adams)
#not surprising that high visits don't statistically predict temp. 

```


# Please include additional sections as desired.

```{r}

```


# Versions

Include some information about the version of R and the packages you are using by running sessionInfo().

```{r}
sessionInfo()
```

Render your document to an html or pdf using the "Knit" button.
